---
title: "Overview of firstapiR Functions"
author: "Stacy Irwin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The _Getting Started_ vignette covered how to use the `GetSession()` function
to create a Session object and how to use that Session object in other
firstapiR functions. This vignette provides an overview of the remaining
firstapiR functions. For detailed information on each function, type
`help(function_name)` at the console or in the RStudio help view search box.


## Simple Functions

There are several functions that take no arguments other than the `Session`
argument.

`GetServerStatus(session)`
  ~ Retrieves information about the FIRST API server.
  
`GetSeason(session)`
  ~ Gets information about the season specified in session$season.

`GetDistricts(session)`
  ~ Provides a list of FIRST districts.
  
`GetAwardsList(session)`
  ~ Obtains a list and description of awards available to FRC teams.
  
  
## Functions with Additional Arguments

All other firstapiR functions take at least a few additional arguments. Most of 
the additional arguments are optional and, usually, only one or two of the
additional arguments are used at any one time.

`GetEvents(session, event, team, district, exclude_district)`
  ~ Provide information about FRC competition events.
  
`GetTeams(session, team, event, district, state, page)`
  ~ Gets information about FIRST FRC teams.
  
`GetSchedule(session, event, level, team, start, end, expand_cols)`
  ~ Downloads the match schedule for a specific event.
  
`GetHybridSchedule(session, event, level, start, end, expand_cols)`
  ~ Downloads the match schedule, along with results for completed matches.
  
`GetMatchResults(session, event, level, team, match, start, end, expand_cols)`
  ~ Gets the scores for the selected matches.
  
`GetScores(session, event, level, team, match, start, end)`
  ~ Downloads the scoring sheet for each match.
  
`GetAlliances(session, event)`
  ~ Obtains the playoff alliances.
  
`GetRankings(session, event, team, top)`
  ~ Gets the team rankings.
  
`GetAwards(session, event, team)`
  ~ Lists the awards and recipients for an FRC competition.


## Function Arguments

Many of the same arguments are used in several different firstapiR functions.

### team

The four digit FRC team number. 

### districts

A code representing a FIRST district. Use GetDistricts() to obtain the district
codes that can be used in other firstapiR functions:

``` {r districtCodes}
library(firstapiR)
sn <- GetSession("username", "key")
districts <- GetDistricts(sn)
```

### event

An event code, such as _WAAMV_ (disctrict competition held at Auburn MountainView
High School in Washington State) or _PNCMP_ (Pacific Northwest District
Championships). Use the GetEvents() functio to get the event codes:

``` {r eventCodes}
sn <- GetSession("username", "key")
PNW_events <- GetEvents(sn, district = "PNW")
PNW_events[, c("code", "name")]
PNW_events[, c("code", "dateStart")]
```

### level

Either "playoff" or "qual". Defaults to "qual".

### start, end, match

Integers specifying the range of matches that will be included in the returned
data.


## Expanded Columns Option

There are three functions that return data about competition matches and that
accept the `expand_cols` argument:

* GetSchedule()
* GetHybridSchedule()
* GetMatchResults()

By default, `expand_cols` is _FALSE_. This causes the function to return a data 
frame with six rows for every match, with one column for team number, one column
for the alliance, etc. This format is preferable for analyzing data. For
example, the following code extracts the top eight teams from the data frame by
average match score and draws a boxplot for each team.

``` {r}
mr <- firstapiR::GetMatchResults(sn, "PNCMP")

# Put the teams in order by average score
mr$teamNumber <- reorder(mr$teamNumber, -mr$scoreFinal, mean)

# Filter out all but the top 8 teams
top8 <- mr$teamNumber[1:8]
mr8 <- mr[mr$teamNumber %in% top8, ]
mr8$teamNumber <- droplevels(mr8$teamNumber)

# Boxplot the top 8 teams
boxplot(scoreFinal~teamNumber, mr8, las = 2, main = "Qual Scores, Top 8 Teams",
        sub = "2016 PNW District Championships")
```

Aggregating team performance data would be more difficult if the scores were
spread across six different columns instead of contained within a single column.

Setting `expand_cols` to _TRUE_ puts all data for the match into a single row
in the data frame with six different columns for teamNumber, surrogate, final
score, etc. This format is useful for displaying an easy to read table:

``` {r}
sched <- firstapiR::GetSchedule(sn, event = "ORPHI", expand_cols = TRUE)

# Display the first 12 rows of the data frame, skipping the first three colums
sched[1:3, 4:17]
```

## Pages Argument

There is one firstapiR function, `GetTeams()`, that accepts the `page`
argument. When requesting data on FRC teams, the FIRST API will split the
results into multiple pages if there are more than 65 teams in the response,
requiring a separate HTTP request for each page.
Users that are requesting the data frame format can ignore the `page` argument
because `GetTeams` will automatically detect if there are multiple pages in the
response, conduct a separate HTTP request for each page, and merge the results
into a single data frame. This feature is not available when XML or JSON formats
are sepecifed in the `Session$format` parameter. Users requesting XML or JSON
formatted data will have to call GetTeams() for each page of data, incrementing
the `page` argument for each request.

## Modified-Since Arguments

There are two additional arguments that can be passed to almost all firstapiR
functions (only `GetSession()` and `GetServerStatus()` do not accept them). The
two arguments, which are always optional, are `mod_since` and `only_mod_since`.
These two arguments help to reduce the load on the FIRST API server by allowing
the server to skip the database query and send only a short HTTP response when
the FIRST API server data has not changed since the user last queried the
server.

Both of these arguments accept a character vector containing an HTTP formatted 
date and time string. If the user includes the `mod_since` argument, and no data
has changed on the FIRST API server since the date and time specified in the 
argument, the FIRST API server will provide an HTTP 304 response with no 
content. When this happens, irstapiR functions will return a logical character
vector with the value `NA`. Here is an example:

```{r eval = FALSE}
# Create an HTTP date-time string set to midnight GMT yesterday
mod_date <- httr::http_date(as.POSIXct(Sys.Date() - 1))

# Request recently changed data from the server
match_results <- firstapiR::GetMatchResults(sn, event = "PNCMP",
                                            mod_since = mod_date)

# Assuming there have been no updates to the data since yesterday, this
#   returns TRUE
is.na(match_results)

# The value passed to mod_since returned as an attribute, even when the
#   result is NA
print(attr(match_results, "mod_since"))

```

If the user includes the `only_mod_since` argument, the FIRST API server will
return only the data that has changed since the date and time specified in the
`only_mod_since` argument. In no changes have been made, the FIRST API server
provides a 304 response and firstapiR functions return a logical vector set to
`NA`.

```{r eval = FALSE}
# Create an HTTP date-time string set to midnight GMT yesterday
mod_date <- httr::http_date(as.POSIXct(Sys.Date() - 1))

# Request recently changed data from the server
match_results <- firstapiR::GetMatchResults(sn, event = "PNCMP",
                                            only_mod_since = mod_date)

# Assuming there have been no updates to the data since yesterday, this
#   returns TRUE
is.na(match_results)

# The value passed to mod_since returned as an attribute, even when the
#   result is NA
print(attr(match_results, "only_mod_since"))

```

If the `mod_since` or `only_mod_since` arguments are not specified, the
corresponding `mod_since` or only_mod_since` attribute attached to the return
value will be set to `NULL`.

Users can keep track of the date and time provided in the `last_modified`
attribute, which corresponds to the most recent date and time that the
information changed on the FIRST API server, and use this date and time to
request only new data from the server. If a user repeats a firstapiR function
call with the same arguments and sets the `mod_since` or `only_mod_since`
argument to the `last_modified` attribute value from the intial firstapiR
function call, the FIRST API server will return all requested data -- the same
data that was provided in the first function call. Users who would rather
receive a 304 response in this situation should either add at least a second to
the `last_modified` value, or should instead use the `time_downloaded` attribute
value to set the `mod_since` or `only_mod_since` arguments.
