---
title: "Overview of firstapiR Functions"
author: "Stacy Irwin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The [Getting Started](http://irwinsnet.github.io/firstapiR/Getting_Started)
vignette covered how to use the `GetSession()` function to create a Session
object and how to use that Session object in other firstapiR functions. This
vignette provides an overview of the remaining firstapiR functions. For detailed
information on each function, type `help(function_name)` at the console or in
the RStudio help view search box.


## Simple Functions for Retrieving FIRST Data

There are several functions that take no arguments other than the `Session`
argument.

`GetServerStatus(session)`
  ~ Retrieves information about the FIRST API server.
  
`GetSeason(session)`
  ~ Gets information about the season specified in session$season.

`GetDistricts(session)`
  ~ Provides a list of FIRST districts.
  
`GetAwardsList(session)`
  ~ Obtains a list and description of awards available to FRC teams.
  
  
## Data Retrieval Functions with Additional Arguments

All other firstapiR functions take at least a few additional arguments. Most of 
the additional arguments are optional and, usually, only one or two of the
additional arguments are used at any one time.

`GetEvents(session, event, team, district, exclude_district)`
  ~ Provide information about FRC competition events.
  
`GetTeams(session, team, event, district, state, page)`
  ~ Gets information about FIRST FRC teams.
  
`GetSchedule(session, event, level, team, start, end)
  ~ Downloads the match schedule for a specific event.
  
`GetHybridSchedule(session, event, level, start, end)`
  ~ Downloads the match schedule, along with results for completed matches.
  
`GetMatchResults(session, event, level, team, match, start, end)`
  ~ Gets the scores for the selected matches.
  
`GetScores(session, event, level, team, match, start, end)`
  ~ Downloads the scoring sheet for each match.
  
`GetAlliances(session, event)`
  ~ Obtains the playoff alliances.
  
`GetRankings(session, event, team, top)`
  ~ Gets the team rankings.
  
`GetAwards(session, event, team)`
  ~ Lists the awards and recipients for an FRC competition.


## Function Arguments

Many of the same arguments are used in several different firstapiR functions.

### team

The four digit FRC team number. 

### districts

A code representing a FIRST district. Use GetDistricts() to obtain the district
codes that can be used in other firstapiR functions:

``` {r districtCodes}
library(firstapiR)
sn <- GetSession("username", "key")
districts <- GetDistricts(sn)
```

### event

An event code, such as _WAAMV_ (disctrict competition held at Auburn MountainView
High School in Washington State) or _PNCMP_ (Pacific Northwest District
Championships). Use the GetEvents() function to get the event codes:

``` {r eventCodes}
sn <- GetSession("username", "key")
PNW_events <- GetEvents(sn, district = "PNW")
PNW_events[, c("code", "name")]
PNW_events[, c("code", "dateStart")]
```

### level

Either "playoff" or "qual". Defaults to "qual".

### start, end, match

Integers specifying the range of matches that will be included in the returned
data.

## Pages Argument

There is one firstapiR function, `GetTeams()`, that accepts the `page`
argument. When requesting data on FRC teams, the FIRST API will split the
results into multiple pages if there are more than 65 teams in the response,
requiring a separate HTTP request for each page.
Users that are requesting the data frame format can ignore the `page` argument
because `GetTeams` will automatically detect if there are multiple pages in the
response, conduct a separate HTTP request for each page, and merge the results
into a single data frame. This feature is not available when XML or JSON formats
are sepecifed in the `Session$format` parameter. Users requesting XML or JSON
formatted data will have to call GetTeams() for each page of data, incrementing
the `page` argument for each request.

## Modified-Since Arguments

There are two additional arguments that can be passed to almost all firstapiR
functions (only `GetSession()` and `GetServerStatus()` do not accept them). The
two arguments, which are always optional, are `mod_since` and `only_mod_since`.
These two arguments help to reduce the load on the FIRST API server by allowing
the server to skip the database query and send only a short HTTP response when
the FIRST API server data has not changed since the user last queried the
server.

Both of these arguments accept a character vector containing an HTTP formatted 
date and time string. If the user includes the `mod_since` argument, and no data
has changed on the FIRST API server since the date and time specified in the 
argument, the FIRST API server will provide an HTTP 304 response with no 
content. When this happens, irstapiR functions will return a logical character
vector with the value `NA`. Here is an example:

```{r eval = FALSE}
# Create an HTTP date-time string set to midnight GMT yesterday
mod_date <- httr::http_date(as.POSIXct(Sys.Date() - 1))

# Request recently changed data from the server
match_results <- firstapiR::GetMatchResults(sn, event = "PNCMP",
                                            mod_since = mod_date)

# Assuming there have been no updates to the data since yesterday, this
#   returns TRUE
is.na(match_results)

# The value passed to mod_since returned as an attribute, even when the
#   result is NA
print(attr(match_results, "mod_since"))

```

If the user includes the `only_mod_since` argument, the FIRST API server will
return only the data that has changed since the date and time specified in the
`only_mod_since` argument. In no changes have been made, the FIRST API server
provides a 304 response and firstapiR functions return a logical vector set to
`NA`.

```{r eval = FALSE}
# Create an HTTP date-time string set to midnight GMT yesterday
mod_date <- httr::http_date(as.POSIXct(Sys.Date() - 1))

# Request recently changed data from the server
match_results <- firstapiR::GetMatchResults(sn, event = "PNCMP",
                                            only_mod_since = mod_date)

# Assuming there have been no updates to the data since yesterday, this
#   returns TRUE
is.na(match_results)

# The value passed to mod_since returned as an attribute, even when the
#   result is NA
print(attr(match_results, "only_mod_since"))

```

If the `mod_since` or `only_mod_since` arguments are not specified, the
corresponding `mod_since` or only_mod_since` attribute attached to the return
value will be set to `NULL`.

Users can keep track of the date and time provided in the `last_modified`
attribute, which corresponds to the most recent date and time that the
information changed on the FIRST API server, and use this date and time to
request only new data from the server. If a user repeats a firstapiR function
call with the same arguments and sets the `mod_since` or `only_mod_since`
argument to the `last_modified` attribute value from the intial firstapiR
function call, the FIRST API server will return all requested data -- the same
data that was provided in the first function call. Users who would rather
receive a 304 response in this situation should either add at least a second to
the `last_modified` value, or should instead use the `time_downloaded` attribute
value to set the `mod_since` or `only_mod_since` arguments.

## Data Frame Shaping Functions

### Team Shape

The functions `GetSchedule()`, `GetHybridSchedule()`, and `GetMatchResults()`
return data frames with one row per team. For example, let's take a look at the
first two qualification matches from the 2016 Pacific Northwest District
Championships:
```{r}
sn <- firstapiR::GetSession("username", "key")
match_results <- firstapiR::GetMatchResults(sn, "PNCMP", level = "qual")
match_results[1:12, c("match", "alliance", "station", "team", "scoreAuto",
                      "scoreFinal")]
```
There is one team listed in each row, and it takes six rows to provide all of
the information from a single match. This is called *team* shape. This format
is useful for calculating summary statistics for individual teams. For example,
to see average scores for the ten teams with the highest scores:
```{r}
agg_scores <- aggregate(match_results[c("scoreFinal", "scoreAuto",
                                        "scoreFoul")],
                        list("team" = match_results$team), mean)
agg_scores[order(agg_scores$scoreFinal, decreasing = TRUE), ][1:10, ]

```
Data frames in team shape will have their *shape* attribute set to "team".
```{r}
attr(match_results, "shape")
```


### Alliance Shape

For some calculations, such as offensive power rating (OPR), the alliance shape,
with three teams per row and two rows per match (one row each for the red and
blue alliances), is more suitable. The function `ToAllianceShape()` will
reshape *Schedule*, *HybridSchedule*, or *MatchResults* data frames from *team*
to *alliance* shape.
```{r}
mresults_alliance <- ToAllianceShape(match_results)
mresults_alliance[1:4, c("match", "alliance", "team.1", "team.2", "team.3",
                         "scoreAuto", "scoreFinal")]
```
Data frames in alliance shape will have their *shape* attribute set to
"alliance".
```{r}
attr(mresults_alliance, "shape")
```

### Match Shape

Finally, *Schedule*, *HybridSchedule*, and *MatchResults* data frames may be
converted from *team* to *match* shape.

```{r}
mresults_match <- ToMatchShape(match_results)
mresults_match[1:2, c("match", "team.Red1", "team.Red2", "team.Red3",
                      "team.Blue1", "team.Blue2", "team.Blue3",
                      "scoreFinal.Blue", "scoreFinal.Red")]
```

### Converting Back to Team Shape

The `ToTeamShape()` function will convert *match* and *alliance* data frames
back to *team* shape. This function depends on the *ReshapeWide* attribute,
which is added to the *match* and *alliance* data frames by R's `reshape()`
function. `ToTeamShape()` will not work if the *ReshapeWide* attribute is
deleted.

### Converting Between Match and Alliance Shape

To convert a data frame from *match* to *alliance* shape or vice-versa, first
convert the data frame back to team shape.

## Merging Team and Score Data

The `GetScores()` function retrieves detailed performance data for for both the
red and blue alliances for each match. However the *Scores* data frame does not
list the teams that were assigned to each alliance. The function `MergeResults()`
merges a *HybridResults* data frame with a *Scores* data frame to create a
*Results* data frame that contains both the detailed match performance data and
the teams that were assigned to each match.

## Merging Qualification and Playoff Data Frames

There are several functions provide data for either qualification or playoff
matches, but not both. Use R's `rbind()` function to merge this data into a
single data frame.

## Saving and Loading Data

Use firstapiR's `SaveData()` function to open a save file dialog box and save
data to an RDS data file. The `ReadData()` function opens an open file dialog
and returns the value saved in the RDS file. Refer to R's documentation on the
`saveRDS()` and `readRDS()' functions for more information on RDS data files.
```{r eval = FALSE}
SaveData(GetAlliances(sn, event = "WAAMV"))
evt_alliances <- ReadData()

```

## Getting all Event Data

Sometimes it's handy to get all the available information for an FRC competition
with single command. The `GetAll()` function retrieves all data for a specific
event and stores the data in a list. I recommend not using this function during
competitions to avoid overloading the server. Also, save the event data to a
file so it only needs to be downloaded once.

## Ask Questions
Post an issue on the [firstapiR github repository](https://github.com/irwinsnet/firstapiR) if you have questions, find bugs or have recommendations for the next version.
