---
title: "Getting Started with firstapiR"
author: "Stacy Irwin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: > 
  %\VignetteIndexEntry{Getting Started} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The firstapiR package provides R functions that download FIRST Robotics 
Competition (FRC) data from the FIRST API server via hypertext transfer protocol
(HTTP) requests. The firstapiR functions format the data as JSON, XML, or as R
data frames.

For obvious reasons, firstapiR should never ever be referred to as the FIRST API
R Toolbox.

### FIRST - *F*or *I*nspiration and *R*ecognition in *S*cience and *T*echnology

FIRST engages kids in kindergarten through high school in exciting,
mentor-based, research and robotics programs that help them become science and
technology leaders, as well as well-rounded contributors to society. Learn more
about FIRST by [visiting their website.](http://www.firstinspires.org/) FRC is
one of four FIRST programs.

In FRC, high-school students, in cooperation with adult mentors,
design and build robots and operate them in competitions. The competitions have
detailed rules ([check out the 2016 game manual ](http://www.firstinspires.org/resource-library/frc/competition-manual-qa-system))
and an extensive amount of data is generated at each competition. All
scheduling, team, scoring, and ranking data is stored in FIRST servers and is
made available to users via the FIRST API.

### Getting Access to the FIRST API

To see what the FIRST API can do, visit the [FIRST API documentation page
](http://docs.frcevents2.apiary.io/#). The documentation page provides detailed
information on the HTTP request formats and authorization headers, as well as
detailed descriptions of the data that is provided by each type of API call.

To retrieve data from the FIRST API via firatapiR functions or via HTTP requests
that you build yourself, you must have a username and authorization key issued
by FIRST. To get a username and authorization key, go to the [FIRST Community
Developers page on TeamForge
](https://usfirst.collab.net/sf/projects/first_community_developers/). Follow
the instructions on the page to join the project and FIRST will email you a
username and authorization key. Make sure you keep your authorization key secret
- don't post the key anywhere that is publicly accessible.

### Learning R

R is a functional, free language intended for statistical computing, data 
analysis and plotting. I initially considered building this package in Python, 
but in the end I wanted to learn to use R, so here we are. Compared to other
languages like Java or Python, R does seem a bit quirky at first, but it has
some very useful, built-in features for working with data. R has an extensive
array of free add-in packages for manipulating and visualizing data, plus there
is an excellent and free IDE available for R, RStudio.

If you are new to R, experimenting with firstapiR isn't the worst way to get
started, but there are some resources you should check out:

  1. Download and user the free version of [RStudio](
  https://www.rstudio.com/home/).
  1. To learn R, I recommend starting with the [Quick R Website
  ](http://www.statmethods.net/)
  1. Next, check out the R introduction on the [main R documentation page
  ](https://cran.r-project.org/manuals.html)
  1. Everything written by [Hadley Wickham](http://hadley.nz/) is excellent. His
  [Advanced R website](http://adv-r.had.co.nz/), [R package website
  ](http://r-pkgs.had.co.nz/) and his [testthat
  ](https://cran.r-project.org/web/packages/testthat/index.html) and [devtools
  ](https://cran.r-project.org/web/packages/devtools/index.html) packages were
  essential to the development of the firstapiR package.

## Getting started - Creating a Session Object

Every HTTP request to the FIRST API requires a username, an authorization key, a
four digit year that identifies the season of interest, and a header that
specifies whether the data should be returned as XML or JSON. Having to type all
of these parameters into every function call would make firstapiR very
cumbersome to use, especially from the R console. The firstapiR package solves
this problem with the `GetSession()` function, which returns a Session object
that contains all of the parameters listed above. All other firstapiR functions
require a Session object as the first parameter. Here's an example:

```{r}
# Create a Session object
sn <- firstapiR::GetSession("username", "key")
sn
```

As you can see the Session object is really just an R list. It's also an
an instance of class "Session", which will be useful to users who want to do
some type checking.

In addition to the username and key, the Session object specifies the season,
data format ("data.frame", "xml", or "json") and whether the staging server
should be used instead of the production server. Any of these parameters can be
included as a named argument when calling GetSession(), or they can be modified
later. For example:
```{r}
sn <- firstapiR::GetSession("username", "key", format = "xml")
sn$season <- 2015
sn
```

## Using the Session Object

The following example shows how the session object is passed to a firstapiR
function:
```{r}
# Use the Session object in a firstapiR function
sn <- firstapiR::GetSession("username", "key")
dist <- firstapiR::GetDistricts(sn)
class(dist)
dist
```

By default, GetSession creates a Session object with the current year, `format =
"data.frame"`, and staging = `TRUE`. Consequently, GetDistricts() returns a data
frame that is also an instance of class "Districts". GetDistricts() is usefule
because the three-letter district codes are used as arguments to several other
firatapiR functions.

## Extracting Data from a Data Frame

R provides numerous methods for extracting, filtering, and manipulating data in
data frames. Refer to the R resources listed in this vignette for additional
information. Individual data elements can be extracted from a data frame using
the format `data_frame$column_name[[row_number]]`. For example:
```{r}
dist$code[[6]]
```

Here's one of many slick techniques:
```{r}
dist$name[dist$code == "PNW"]
```

## Why These Examples Work: Local verses HTTP Data

Hopefully you were able to figure out that "username" and "key" are not
afilliated with a valid FIRST API account. These examples work because
for all firstapiR functions, if the key value in the Session object
is the literal value "key", firstapiR skips the HTTP requests and extracts
cached JSON or XML data from the R/sysdata.rda file. If you insert any other
random value into the key parameter, or the username does not match an acutal
FIRST API username, the the FIRST API will return a 401 status code and
the firstapiR function will throw an error.

All of the firstapirR functions add attributes to whatever data they return. 
These attributes allow you to identify whether the data is local or was
extracted directly from the FIRST API server. To demonstrate, we'll analyze the
Districts data frame using R's str() function, which displays the structure of
any object:

```{r}
str(dist)
```

The attribute "local_test_data" is set to TRUE if the object was created from
cached data in the R/sysdata.rda file. The "local_url" contains the actual URL
that was sent to the FIRST API server to request the data, and the
"time_downloaded" shows when the data was downloaded from the server.

For non-local, HTTP data, "local_test_data" will be set to FALSE, and
"local_url" will be NULL. The "time_downloaded" attribute will contain the time
that the HTTP request was received, based on the local system time.

There is also a "url" attribute. This attribute always contains the URL that is
constructed by firstapiR based on the arguments passed to the firstapiR
function. Normally, this is the actual URL that is sent to the FIRST API server,
but when extracting local data, this URL is mostly ignored. In the
GetDistricts() example, "local_url" matches "url" because GetDistricts has no
optional parameters and there is only one way to call this function. Most other
firstapiR functions have many optional parameters and a multitude of possible
return values. Since it's not practical to cache the entire FIRST API server,
when requesting local data (by setting the key parameter to "key"), firstapiR
ignores the optional parameters. The local data will be of the same type and
format as what was requested. In these situations the "url" attribute will
represent was was requested, and the "local_url" attribute will describe what
was cached and returned.
