---
title: "Getting Started with firstapiR"
author: "Stacy Irwin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: > 
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

The firstapiR package provides R functions that download FIRST Robotics 
Competition (FRC) data from the FIRST API server via hypertext transfer protocol
(HTTP) requests. The firstapiR package formats the resulting data as JSON text,
XML text, or as R data frames.


### Version 2.0.0

This guide is for version 2.0.0 of the firstapiR package. [Guidance for version
1.0.0 of firstapiR is located here
](http://irwinsnet.github.io/projects/firstapiR/Getting_Started_v100).


### FIRST - *F*or *I*nspiration and *R*ecognition in *S*cience and *T*echnology

FIRST engages kids in kindergarten through high school in exciting, 
mentor-based, research and robotics programs that help them become science and 
technology leaders, as well as well-rounded contributors to society. Learn more 
about FIRST by [visiting their website.](http://www.firstinspires.org/)

FRC is one of four FIRST programs. In FRC, high-school students, in cooperation
with adult mentors, design and build robots and operate them in competitions.
The competitions have detailed rules ([check out the 2016 game manual 
](http://www.firstinspires.org/resource-library/frc/competition-manual-qa-system))
and an extensive amount of data is generated at each competition. All 
scheduling, team, scoring, and ranking data is stored in FIRST servers and is 
made available to users via the FIRST API.


### Getting Access to the FIRST API

To see what the FIRST API can do, visit the [FIRST API documentation page 
](http://docs.frcevents2.apiary.io/#). The documentation page provides detailed 
information on the formats for HTTP request and authorization headers, as well
as detailed descriptions of the data provided by each type of API call.

To use the FIRST API (via firstapiR functions or via HTTP requests that you
build yourself), you must have a username and authorization key issued by FIRST.
To get a username and authorization key, go to the [FIRST Community Developers
page on TeamForge 
](https://usfirst.collab.net/sf/projects/first_community_developers/). Follow 
the instructions on the page to join the project and FIRST will email you a 
username and authorization key. Make sure you keep your authorization key secret
- don't post the key anywhere that is publicly accessible.


### Learning R

R is a functional language intended for statistical computing, data analysis and
plotting. I initially considered building this package in Python, but in the end
I wanted to learn to use R, so here we are. R seems a bit quirky at first, 
especially compared to other languages like Java or Python, but it has some
useful, built-in features for working with data. R has an extensive array of
free add-in packages for manipulating and visualizing data, plus there is an
excellent and free IDE available for R, RStudio.

If you are new to R, experimenting with firstapiR isn't the worst way to get 
started, but there are some resources you should check out:

1. Download R from [the main R
website](https://cran.r-project.org/mirrors.html).
1. Download and use the free version of [RStudio]( 
https://www.rstudio.com/home/).
1. To learn R, I recommend starting with the
[Quick R Website ](http://www.statmethods.net/)
1. Next, check out the R
introduction on the [main R documentation page 
](https://cran.r-project.org/manuals.html)
1. Everything written by [Hadley
Wickham](http://hadley.nz/) is excellent. His [Advanced R
website](http://adv-r.had.co.nz/), [R package website 
](http://r-pkgs.had.co.nz/) and his [testthat 
](https://cran.r-project.org/web/packages/testthat/index.html) and [devtools 
](https://cran.r-project.org/web/packages/devtools/index.html) packages were 
essential to the development of the firstapiR package.


### Installing the firstapiR package

1. Install the devtools package. This is easy to do in RStudio - just select
_Tools->Install Packages..._ from the menu bar, type "devtools" in the
package name box and hit _Install_.

1. The devtools package provides a function for installing R packages from
GitHub. Type the following command at the RStudio console prompt:
```{r install_v2.0.0, eval = FALSE}
devtools::install_github("irwinsnet/firstapiR")
```

If you need the prior version of firstapiR, use this command instead:
```{r install_v1.0.0, eval= FALSE}
devtools::install_github("irwinsnet/firstapiR", ref = "v1.0.0")
```

You can also download source or binary versions of the firstapiR package
[from this website](https://irwinsnet.github.io/). Type
`help(install.packages)` in the R or Rstudio console to see the documentation
for installing packages.

## Differences between Versions 1.0.0 and 2.0.0

1. Version 1.0.0 of firstapiR is still available from github per the firstapiR
installation instructions above.

1. The main difference from version 1.0.0 is that there are three shapes (i.e, 
types) of *Schedule*, *HybridSchedule*, and *MatchResult* data frames. The three
shapes are *team*, *alliance*, and *match*. Team data frames have one row per
team, requiring six rows to contain all of the data for a single match. The 
*station* column specifies the station to which the team was assigned. Alliance 
data frames have three teams per row, with one row for the blue alliance and 
another for the red alliance. The *alliance* column contains *Red* or *Blue*,
depending on the alliance. Finally, match data frames have six teams per row
and one row per match.

1. The functions `GetSchedule()`, `GetHybridSchedule()`, and `GetMatchResults()`
no longer accept the `expand_cols` argument. These functions can only return a
data frame in team shape. Users can convert between team, alliance, and match 
data frames with the functions `ToAllianceShape()`, `ToMatchShape()`, and 
`ToTeamShape()`.

1. I've added a `MergeReults()` function which merges the *HybridSchedule* and
*Scores* data frames into a single data frame, providing team numbers and
detailed performance data all in one table.

1. I've added the function `GetAll()`, which downloads everything about a single
competition and saves it in a single list.

1. I've added functions for saving FIRST competition data locally in R data
files. The functions `SaveData()` and `ReadData()` display file open and save
dialogs for saving and loading data to and from *.RDS* files.

1. To give users without an authorization key some data to play with, firstapiR
now includes all performance data for the 2016 FIRST Robotics Championships.
The data is available in R lists named `archimedes2016`, `carver2016`,
`curie2016`, etc.

1. I've fixed a few inconsistencies. Some character columns were changed to
factors, some column names were shortened to reduce excess typing, row names are
more consistent, etc.

## Getting Started - Creating a Session Object

Every HTTP request to the FIRST API requires a username, an authorization key, a
four digit year that identifies the season of interest, and a header that 
specifies whether the data should be returned as XML or JSON. Having to type all
of these parameters into every function call would make firstapiR very 
cumbersome to use, especially from the R console. The firstapiR package
addresses this problem with the `GetSession()` function, which returns a Session
object that contains all of the parameters listed above. All other firstapiR
functions require a Session object as the first parameter. Here's an example:

```{r GetSession_1}
# Create a Session object
sn <- firstapiR::GetSession("username", "key")

# Display the Session object.
sn
```

As you can see the Session object is really just an R list. It's also an 
instance of class "Session", which will be useful to users who want to do some 
type checking.

In addition to the username and key, the Session object specifies the season, 
data format ("data.frame", "xml", or "json") and whether the staging server 
should be used instead of the production server. Any of these parameters can be 
included as a named argument when calling GetSession(), or they can be modified 
later. For example:

```{r GetSession_2}
# Create a Session and specify XML format
sn <- firstapiR::GetSession("username", "key", format = "xml")

# Change the season to 2015
sn$season <- 2015

# Display the Session object
sn
```


## Using the Session Object

The following example shows how the session object is passed to a firstapiR 
function:

```{r Districts}
# Create a Session and use it in a firstapiR function
sn <- firstapiR::GetSession("username", "key")
dist <- firstapiR::GetDistricts(sn)

# Display the class of the object returned by GetDistricts()
class(dist)

# Display the data frame
dist
```

By default, GetSession creates a Session object with the current year, `format =
"data.frame"`, and staging = `TRUE`. `GetDistricts()` returns a data frame that
is also an instance of class "Districts". `GetDistricts()` is useful because the
three-letter district codes are used as arguments to several other firstapiR 
functions.

**Note**: If you run the example above, passing the string "key" to 
GetSession(), firstapiR skips the HTTP request and pulls data from a local
cache. See _Why These Examples Work: Local verses HTTP Data_ below for an
explanation of this feature.


## Extracting Data from a Data Frame

R provides numerous methods for extracting, filtering, and manipulating data in 
data frames. Refer to the R resources listed in this vignette for additional 
information. Individual data elements can be extracted from a data frame using 
the syntax `data_frame$column_name[[row_number]]`. For example:

```{r Subsetting_1}
# Retrieve the item in th 6th row of the column named "code"
dist$code[[6]]
```

Here's one of many slick techniques that are available in R:

```{r Subsetting_2}
# Retrieve the name of any district with code == "PNW"
dist$name[dist$code == "PNW"]
```


## Next Steps

This package provides a detailed help file for every firstapiR function. Make 
sure the firstapiR package is loaded by running the command 
`library(firstapiR)`. Then in RStudio, just type the name of the function in the
help view search box. Or type `help(function_name)` at the R console command
prompt. Here are the firstapiR functions:

|                   |                   |                |                    | 
|:------------------|:------------------|:---------------|:-------------------|
| GetSession()      | GetServerStatus() | GetSeason()    | GetDistricts()     |
| GetEvents()       | GetTeams()        | GetSchedule()  | GetHybridSchedule()|
| GetMatchResults() | GetScores()       | GetAlliances() | GetRankings()      |
| GetAwards()       | GetAwardsList()   | GetAll()       | MergeResults()     |
| ReadData()        | SaveData()        | ToTeamShape()  | ToMatchShape()     |
| ToAllianceShape() |

Table: **External Functions in firstapiR Package**

In addition to help documents on each function, this package contains one other
vignette: 
[Overview of firstapiR Functions](http://irwinsnet.github.io/firstapiR/Functions).


## Advanced Topics

### Why These Examples Work: Local verses HTTP Data

Hopefully you were able to figure out that "username" and "key" are not 
affiliated with a valid FIRST API account. These examples work because for all
firstapiR functions, if the key value in the Session object is the literal value
"key", firstapiR skips the HTTP requests and extracts cached JSON or XML data
from the R/sysdata.rda file. If you insert any other random value into the key
parameter, or the username does not match an actual FIRST API username, the
FIRST API will return a 401 status code and the firstapiR function will throw an
error.

I added the local data feature for several reasons.

* Users can experiment with firstapiR functions and the data they provide 
without a FIRST API user account.
* The examples in the help documents and vignettes will work exactly as typed.
* Most of firstapiR's functionality can be tested without internet access or a
FIRST API account.


### Attributes of Returned firstapiR Objects

All of the firstapiR functions add attributes to whatever data they return. 
These attributes allow you to identify whether the data is local or was 
extracted directly from the FIRST API server, when the data was extracted from 
the FIRST API server, and the URL that was sent to the server. To demonstrate, 
we'll analyze the Districts data frame using R's str() function, which displays
the structure of any object:

```{r structure}
# Display the structure and attributes of the dist object
str(dist)
```

The attribute _local_test_data_ is set to TRUE if the object was created from 
cached data in the R/sysdata.rda file. The _local_url_ attribute contains the
actual URL that was sent to the FIRST API server to request the data, and the 
_time_downloaded_ attribute shows when the data was downloaded from the server.

For non-local, HTTP data, _local_test_data_ will be set to FALSE, and 
_local_url_ will be NULL. The _time_downloaded_ attribute will contain the time 
that the HTTP request was received, based on the local system time.

There is also a _url_ attribute. This attribute always contains the URL that is 
constructed from the arguments passed to the firstapiR function. Normally, this
is the actual URL that is sent to the FIRST API server, but when extracting
local data, this URL is mostly ignored. In the GetDistricts() example,
_local_url_ matches _url_ because GetDistricts has no optional parameters and
there is only one way to call this function. Most other firstapiR functions have
many optional parameters and a multitude of possible return values. Since it's
not practical to cache the entire FIRST API server, when requesting local data
(by setting the key parameter to "key"), firstapiR ignores the optional
parameters. The local data will be of the same type and format as what was
requested. In these situations the _url_ attribute will represent was was
requested, and the _local_url_ attribute will describe what was cached and
returned.

Use the `attr(object, attribute_name)` function and syntax to retrive an
attribute from an object.

Open the _tools/data.R_ folder (only available in the firstapiR github 
repository) to see the firstapiR commands that were used to generate the local
data.


### Testing firstapiR on Your System

The firstapiR package includes several automated tests in the tests/testthat 
folder. Users can run these tests if they have the _testthat_ package installed 
on their system. Run all of the tests by running 
`testthat::test_dir(path_to_tests/testthat_folder)` at the R console.

If you run the `test_dir` command, you'll probably notice that several tests are
skipped. The tests are set up such that most of them can run without an internet
connection or a FIRST API account. This is OK because local data requests use
almost all of the firstapiR code that would be used for an actual HTTP request,
including evaluation of the function arguments, constructing the URL and HTTP
headers, and creating and formatting the data frame from JSON text. (If you
specify the data frame format, firstapiR actually requests JSON text from the
FIRST API server, because JSON is more compact than XML.) Only the actual HTTP
request is skipped.

If you have a username and authorization key assigned by FIRST, then you can run
all tests. Store your username and key in character vectors named "username" and
"key". Then run `test_dir` from the R console. The tests will detect that the 
variables username and key exist and will run tests that make actual HTTP calls 
to the FIRST API server. If either your username or key are incorrect, the tests 
will fail due to an HTTP 401 error.


## Ask Questions
Post an issue on the [firstapiR github
repository](https://github.com/irwinsnet/firstapiR) if you have questions, find
bugs or have recommendations for the next version.

## Next Vignette

[Overview of firstapiR Functions](http://irwinsnet.github.io/firstapiR/Functions)

## Links

[Irwinsnet Home Page](http://irwinsnet.github.io)
[firstapiR Local Data](http://irwinsnet.github.io/firstapiR/Data)
